Testing Quasar Unification: I. Radiative Transfer in Clumpy Winds

### 2015-12-23 paper.tex
slimming down

\documentclass[useAMS,usenatbib]{mn2e_x}
%%%% Various options for document class.
%\usepackage{psfig,morefloats,url}
%use preprint2 for 2 columns paper.

@@

\title
[Radiative Transfer in Clumpy Winds]{
% Modelling quasar outflows:
% clumpy winds and broad emission lines
Testing Quasar Unification: Radiative Transfer in Clumpy Winds
%Unifying quasars with clumpy wind models
}

@@

\begin{abstract} 
Various unification schemes interpret the complex phenomenology of quasars and luminous active galactic nuclei (AGN) in terms of a simple picture involving a central black hole, an accretion disc and an associated outflow. Here, we continue our tests of this paradigm by comparing quasar spectra to synthetic spectra of biconical disc wind models, produced with our state-of-the-art Monte Carlo radiative transfer code. Previously, we have shown that we could produce synthetic spectra resembling those of observed broad absorption line (BAL) quasars, but only if the X-ray luminosity was limited to
10$^{43}$~erg~s$^{-1}$. Here, we show that clumping the outflow can maintain the ionization state necessary for BAL features in the rest-frame UV at more realistic X-ray luminosities. Our fiducial model shows good agreement with X-ray properties of AGN and the dense, X-ray heated wind produces strong line emission in, e.g., \la\ and \civline at low inclinations. At high inclinations, the synthetic spectra possess prominent LoBAL features. Despite these successes, we cannot reproduce all the emission lines seen in quasar spectra with the correct equivalent-width ratios, and and we find an angular dependence of emission-line equivalent width despite the remarkable similarities in the observed emission line properties of BAL and non-BAL quasars. Overall, our work suggests that a biconical wind can reproduce much of the qualitative behaviour expected from a unified model, but we cannot yet provide quantitative matches with quasar properties at all viewing angles. Whether disc winds can successfully unify quasars is therefore still an open question.
\end{abstract}

@@

The spectra of 
quasars and luminous active galactic nuclei (AGN) 
typically exhibit a series of strong emission lines
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is often attributed to emission from a 
geometrically thin, optically thick accretion disc surrounding the central black hole (BH), similar to that described by \cite{shakurasunyaev1973}.
% {\bf Note: the first half of this paragraph and the second do not follow from one another.  
% The first half is about spectra the second about outflows and the connection to spectra is not made until the next paragrph.  
% This needs fixing. JM: Knox, I like this structure...I start with inflow, then move to outflow.}
In addition to the {\em inflowing} accreting material, 
{\em outflows} are ubiquitous in AGN
and quasars \citep{kellerman1989,ganguly2008}. These outflows can take the form of 
highly collimated radio jets \citep[e.g.][]{hazard1963,potash1980,perley1984,marscher2006}, 
or mass-loaded `winds' emanating from the accretion disc 
\citep{weymann1991,turnermiller2009}. 
Outflows in AGN offer a 
potential feedback mechanism through which the central source can 
affect its environment \citep{king2003,king2005,fabian2012}
-- feedback that is required in models of galaxy evolution \citep{springel2005}
and may explain the `$M-\sigma$' relation \citep{silkrees1998,haring2004}.

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/fig2.eps}
\caption
{
Contour plots showing the logarithm of some important 
physical properties of the outflow. The spatial scales are
logarithmic and the $x$ and $z$ scales are not the same.
Symbols are defined in the text.
The solid black line marks a sphere at $1000~r_G$.
The dotted lines show the $72^\circ$ and $78^\circ$ sightlines 
to the centre of the system, and illustrate that different sightlines
intersect material of different ionization states.
The line luminosities, $L$, represent the luminosity of photons
escaping the Sobolev region for each line. These photons do not
necessarily escape to infinity.
}
\label{fig:wind}
\end{figure} %fullpage

@@

One common way to quantify the ionization state of a plasma
is through the ionization parameter, $U_H$, given by
\begin{equation}
U_H = \frac{4\pi}{n_H c}\int_{13.6{\rm{eV}} / h}^{\infty}\frac{{J_{\nu}d\nu}}{h\nu}.
\end{equation}
\noindent where $\nu$ denotes photon 
frequency. Shown in Fig.~\ref{fig:wind},
the ionization parameter is a useful measure of the global ionization state,
as it represents the ratio of the number density of 
H ionizing photons to the local H density.
It is, however, a poor representation of the 
ionization state of species such as \civ\ as it encodes no information
about the shape of the SED. In our case, the X-ray photons 
are dominant in the photoionization of the UV resonance line ions. 
This explains why a factor of 100 increase in X-ray luminosity requires
a clumping factor of 0.01, even though the value of $U_H$ decreases by only a factor of $\sim10$ compared to H13. 
%This is also the reason for significant Lyman edge photoabsorption
%at the highest inclinations (see section 4.3).

The total line luminosity also increases dramatically compared to the unclumped model
described by H13. This is because the denser outflow can absorb the increased
X-ray luminosity without becoming over-ionized, leading to a hot plasma which
produces strong collisionally excited line emission.
This line emission typically emerges on the edge of the wind
nearest the central source. The location of the line emitting regions
is dependent on the ionization state, as well as the incident X-rays.
The radii of these emitting regions is important,
and can be compared to observations. The line luminosities, $L$,
shown in the figure correspond to the luminosity in $\LUM$ of photons
escaping the Sobolev region for each line. 
% This is equivalent to $\beta_{ul} n_u A_{ul}$, where the three quantities
% represent the Sobolev escape probability, upper level number density and 
% Einstein A coefficient for the line. 
As shown in Fig.~\ref{fig:wind},
the \civline\ line in the fiducial model is typically formed between 
$100-1000~r_G$ ($\sim10^{17}-10^{18}~\rm{cm}$).
This is in rough agreement with the reverberation mapping 
results of Kaspi (2000) for the $2.6\times10^{9} M_\odot$ quasar S5 0836+71,
and also compares favourably with microlensing measurements of the size of the
\civline\ emission line region in the BALQSO H1413+117 \citep{odowd2015}.

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/fig3.eps}
\caption
{
Synthetic spectra at four viewing angles for the fiducial model. At 
$20^\circ$ and $60^\circ$ we show a comparison to an SDSS quasar composite
from Recihard et al. (2003). At $73^\circ$ and $76^\circ$ we show a comparison to
an {\sl HST} STIS spectrum of the high BALnicity BALQSO 
PG0946+301 (Arav et al. 2000), and an SDSS spectrum of the LoBAL quasar 
SDSS J162901.63+453406.0, respectively. The dotted line shows a disc
only continuum to show the effect of the outflow on the continuum level. 
All the spectra are scaled to the model flux at $2000$\AA, expect for the 
{\sl HST} STIS spectrum of PG0946+301, which is scaled to $1350$\AA\
due to the incomplete wavelength coverage.
% HST STIS spectrum of PG0946+301 (Arav et al. 2000)
}
\label{fig:uvspec}
\end{figure} %fullpage

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/fig5.eps}
\caption
{
Synthetic spectra at two viewing angles, 
this time in frequency space and including the optical band,
compared to the non-BAL SDSS quasar composite. The spectra are normalised to the flux at 
$2000$\AA, then an offset of 2 is applied per spectrum for clarity -- the dotted lines show the zero point of $F_\nu / F_{2000\AA}$ in each case.
}
\label{fig:sed}
\end{figure} %fullpage

@@

The high ionization BAL profiles are often saturated, and the location in velocity space
of the strongest absorption in the profile varies with inclination.
At the lowest inclination BAL sight lines, the strongest absorption occurs at the red edge,
whereas at higher inclinations (and for the strongest BALs)
the trough has a sharp edge at the terminal velocity.
This offers one potential explanation for the wide range of BALQSO absorption
line shapes (see e.g. Trump et al. 2006; Knigge et al 2008, Filiz Ak et al. 2014).
% In addition, the line profile shape is strongly dependent 
% on the density, ionization and velocity 
% profiles intersected by the line of sight. Thus, small tweaks of the velocity
% law and angular distributions of streamlines can dramatically alter
% the shape of the line.

The absorption profiles seen in BALQSOs are often non-black, but saturated, 
with flat bases to the absorption troughs \citep{arav1999b,arav1999a}.
%Non-black saturation is observed in the absorption troughs of BALQSOs \citep{arav1999b,arav1999a}.
This is usually explained either as  partial covering of the continuum
source or by scattered contributions to the BAL troughs, necessarily
from an opacity source not co-spatial with the BAL forming region.
The scattered light explanation is supported by spectropolarimetry results
\citep{lamy2000}. Our spectra do not show non-black, saturated profiles.
We find black, saturated troughs at angles $i > 73^\circ$, and the BALs
are non-saturated at lower inclinations. The reasons for this are inherent 
in the construction of our model. 
First, the microclumping assumption does not allow for 
porosity in the wind, meaning that it does not naturally produce
a partial covering absorber. To allow this, an alternative approach
such as {\em macroclumping} would be required \citep[e.g.][]{hamann2008,surlan2012}.
Second, our wind does not have a significant scattering contribution 
along sightlines which do not pass through the BAL region,
meaning that any scattered component to the BAL troughs is absorbed by line opacity.
This suggests that either the scattering cross-section of the wind must
be increased (with higher mass loss rates or covering factors), or 
that an additional source of electron opacity is required, potentially
in a polar direction above the disc. We note the scattering contribution
from plasma in polar regions is significant in some `outflow-from-inflow'
simulations \citep{KP09, simproga2012}.

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=0.9\textwidth]{figures/fig6.eps}
\caption
{
X-ray ($2$~keV) luminosity of the our clumped model (red squares) 
and the H13 model (blue squares), plotted against monochromatic luminosity 
at 2500\AA. The points are labeled according to inclination; angles
$>70^\circ$ correspond to BALs in our scheme (see figure 4).
Also plotted are masurements from 
the COMBI-7 AGN and the BQS samples (Steffen et al. 2006) and the Saez et al. (2012) 
sample of BALQSOs. The dotted line shows the best fit relation for non-BALQSOs 
from Steffen et al. (2006).
}
\label{fig:xray}
\end{figure} %fullpage

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=0.8\textwidth]{figures/fig7.eps}
\caption
{
\civ , \mg , \al\ and Fe~\textsc{ii} line profiles for viewing angles
from $72-79^\circ$. The profiles are plotted relative to the local
continuum with an offset applied for clarity. Lower ionization
profiles appear at a subset of high inclinations, compared
to the ubiquitous \civ\ profile.
}
\label{fig:lobal}
\end{figure} %fullpage

@@

\begin{figure} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/fig8.eps}
\caption
{
The EW of the \civ~$1550$\AA\ line at $20^\circ$ plotted against a) the 
$BI$ of \civ~$1550$\AA\ at $75^\circ$, b) the EW of the \mg~$2800$\AA\ line 
at $20^\circ$ and c) the EW of \la\ at $20^\circ$. The circles correspond 
to the simulation grid for two different values of $f_V$, and the fiducial 
model is marked with an orange star. 
We also show a higher X-ray luminosity model and a higher mass loss rate
with a red star.
}
\label{fig:grid}
\end{figure} %fullpage

@@

\bibliography{mybib.bib,stellar.bib,h14.bib,h13.bib,hamann.bib,krolik.bib}
\clearpage
\clearpage

>>>>

\documentclass[useAMS,usenatbib]{mn2e_x}
%%%% Various options for document class.

@@

\title
[Radiative Transfer in Clumpy Winds]
{
Testing Quasar Unification: Radiative Transfer in Clumpy Winds
}

@@

\maketitle
\begin{abstract} 
Various unification schemes interpret the complex phenomenology of quasars and luminous active galactic nuclei (AGN) in terms of a simple picture involving a central black hole, an accretion disc and an associated outflow. Here, we continue our tests of this paradigm by comparing quasar spectra to synthetic spectra of biconical disc wind models, produced with our state-of-the-art Monte Carlo radiative transfer code. Previously, we have shown that we could produce synthetic spectra resembling those of observed broad absorption line (BAL) quasars, but only if the X-ray luminosity was limited to
10$^{43}$~erg~s$^{-1}$. Here, we show that clumping the outflow can maintain the ionization state necessary for BAL features in the rest-frame UV at more realistic X-ray luminosities. Our fiducial model shows good agreement with X-ray properties of AGN and the dense, X-ray heated wind produces strong line emission in, e.g., \la\ and \civline at low inclinations. At high inclinations, the synthetic spectra possess prominent LoBAL features. Despite these successes, we cannot reproduce all the emission lines seen in quasar spectra with the correct equivalent-width ratios, and and we find an angular dependence of emission-line equivalent width despite the remarkable similarities in the observed emission line properties of BAL and non-BAL quasars. Overall, our work suggests that a biconical wind can reproduce much of the qualitative behaviour expected from a unified model, but we cannot yet provide quantitative matches with quasar properties at all viewing angles. Whether disc winds can successfully unify quasars is therefore still an open question.
\end{abstract}

@@

The spectra of 
quasars and luminous active galactic nuclei (AGN) 
typically exhibit a series of strong emission lines
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is often attributed to emission from a 
geometrically thin, optically thick accretion disc surrounding the central black hole (BH), similar to that described by \cite{shakurasunyaev1973}.
In addition to the {\em inflowing} accreting material, 
{\em outflows} are ubiquitous in AGN
and quasars \citep{kellerman1989,ganguly2008}. These outflows can take the form of 
highly collimated radio jets \citep[e.g.][]{hazard1963,potash1980,perley1984,marscher2006}, 
or mass-loaded `winds' emanating from the accretion disc 
\citep{weymann1991,turnermiller2009}. 
Outflows in AGN offer a 
potential feedback mechanism through which the central source can 
affect its environment \citep{king2003,king2005,fabian2012}
-- feedback that is required in models of galaxy evolution \citep{springel2005}
and may explain the `$M-\sigma$' relation \citep{silkrees1998,haring2004}.

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/fig2.eps}
\caption
{
Contour plots showing the logarithm of some important 
physical properties of the outflow. The spatial scales are
logarithmic and the $x$ and $z$ scales are not the same.
Symbols are defined in the text.
The solid black line marks a sphere at $1000~r_G$.
The dotted lines show the $72^\circ$ and $78^\circ$ sightlines 
to the centre of the system, and illustrate that different sightlines
intersect material of different ionization states.
The line luminosities, $L$, represent the luminosity of photons
escaping the Sobolev region for each line. These photons do not
necessarily escape to infinity.
}
\label{fig:wind}
\end{figure*}

@@

One common way to quantify the ionization state of a plasma
is through the ionization parameter, $U_H$, given by
\begin{equation}
U_H = \frac{4\pi}{n_H c}\int_{13.6{\rm{eV}} / h}^{\infty}\frac{{J_{\nu}d\nu}}{h\nu}.
\end{equation}
\noindent where $\nu$ denotes photon 
frequency. Shown in Fig.~\ref{fig:wind},
the ionization parameter is a useful measure of the global ionization state,
as it represents the ratio of the number density of 
H ionizing photons to the local H density.
It is, however, a poor representation of the 
ionization state of species such as \civ\ as it encodes no information
about the shape of the SED. In our case, the X-ray photons 
are dominant in the photoionization of the UV resonance line ions. 
This explains why a factor of 100 increase in X-ray luminosity requires
a clumping factor of 0.01, even though the value of $U_H$ decreases by only a factor of $\sim10$ compared to H13.

The total line luminosity also increases dramatically compared to the unclumped model
described by H13. This is because the denser outflow can absorb the increased
X-ray luminosity without becoming over-ionized, leading to a hot plasma which
produces strong collisionally excited line emission.
This line emission typically emerges on the edge of the wind
nearest the central source. The location of the line emitting regions
is dependent on the ionization state, as well as the incident X-rays.
The radii of these emitting regions is important,
and can be compared to observations. The line luminosities, $L$,
shown in the figure correspond to the luminosity in $\LUM$ of photons
escaping the Sobolev region for each line. 
As shown in Fig.~\ref{fig:wind},
the \civline\ line in the fiducial model is typically formed between 
$100-1000~r_G$ ($\sim10^{17}-10^{18}~\rm{cm}$).
This is in rough agreement with the reverberation mapping 
results of Kaspi (2000) for the $2.6\times10^{9} M_\odot$ quasar S5 0836+71,
and also compares favourably with microlensing measurements of the size of the
\civline\ emission line region in the BALQSO H1413+117 \citep{odowd2015}.

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/fig3.eps}
\caption
{
Synthetic spectra at four viewing angles for the fiducial model. At 
$20^\circ$ and $60^\circ$ we show a comparison to an SDSS quasar composite
from Recihard et al. (2003). At $73^\circ$ and $76^\circ$ we show a comparison to
an {\sl HST} STIS spectrum of the high BALnicity BALQSO 
PG0946+301 (Arav et al. 2000), and an SDSS spectrum of the LoBAL quasar 
SDSS J162901.63+453406.0, respectively. The dotted line shows a disc
only continuum to show the effect of the outflow on the continuum level. 
All the spectra are scaled to the model flux at $2000$\AA, expect for the 
{\sl HST} STIS spectrum of PG0946+301, which is scaled to $1350$\AA\
due to the incomplete wavelength coverage.
% HST STIS spectrum of PG0946+301 (Arav et al. 2000)
}
\label{fig:uvspec}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/fig5.eps}
\caption
{
Synthetic spectra at two viewing angles, 
this time in frequency space and including the optical band,
compared to the non-BAL SDSS quasar composite. The spectra are normalised to the flux at 
$2000$\AA, then an offset of 2 is applied per spectrum for clarity -- the dotted lines show the zero point of $F_\nu / F_{2000\AA}$ in each case.
}
\label{fig:sed}
\end{figure*}

@@

The high ionization BAL profiles are often saturated, and the location in velocity space
of the strongest absorption in the profile varies with inclination.
At the lowest inclination BAL sight lines, the strongest absorption occurs at the red edge,
whereas at higher inclinations (and for the strongest BALs)
the trough has a sharp edge at the terminal velocity.
This offers one potential explanation for the wide range of BALQSO absorption
line shapes (see e.g. Trump et al. 2006; Knigge et al 2008, Filiz Ak et al. 2014).

The absorption profiles seen in BALQSOs are often non-black, but saturated, 
with flat bases to the absorption troughs \citep{arav1999b,arav1999a}.
This is usually explained either as  partial covering of the continuum
source or by scattered contributions to the BAL troughs, necessarily
from an opacity source not co-spatial with the BAL forming region.
The scattered light explanation is supported by spectropolarimetry results
\citep{lamy2000}. Our spectra do not show non-black, saturated profiles.
We find black, saturated troughs at angles $i > 73^\circ$, and the BALs
are non-saturated at lower inclinations. The reasons for this are inherent 
in the construction of our model. 
First, the microclumping assumption does not allow for 
porosity in the wind, meaning that it does not naturally produce
a partial covering absorber. To allow this, an alternative approach
such as {\em macroclumping} would be required \citep[e.g.][]{hamann2008,surlan2012}.
Second, our wind does not have a significant scattering contribution 
along sightlines which do not pass through the BAL region,
meaning that any scattered component to the BAL troughs is absorbed by line opacity.
This suggests that either the scattering cross-section of the wind must
be increased (with higher mass loss rates or covering factors), or 
that an additional source of electron opacity is required, potentially
in a polar direction above the disc. We note the scattering contribution
from plasma in polar regions is significant in some `outflow-from-inflow'
simulations \citep{KP09, simproga2012}.

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/fig6.eps}
\caption
{
X-ray ($2$~keV) luminosity of the our clumped model (red squares) 
and the H13 model (blue squares), plotted against monochromatic luminosity 
at 2500\AA. The points are labeled according to inclination; angles
$>70^\circ$ correspond to BALs in our scheme (see figure 4).
Also plotted are masurements from 
the COMBI-7 AGN and the BQS samples (Steffen et al. 2006) and the Saez et al. (2012) 
sample of BALQSOs. The dotted line shows the best fit relation for non-BALQSOs 
from Steffen et al. (2006).
}
\label{fig:xray}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=0.8\textwidth]{figures/fig7.eps}
\caption
{
\civ , \mg , \al\ and Fe~\textsc{ii} line profiles for viewing angles
from $72-79^\circ$. The profiles are plotted relative to the local
continuum with an offset applied for clarity. Lower ionization
profiles appear at a subset of high inclinations, compared
to the ubiquitous \civ\ profile.
}
\label{fig:lobal}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/fig8.eps}
\caption
{
The EW of the \civ~$1550$\AA\ line at $20^\circ$ plotted against a) the 
$BI$ of \civ~$1550$\AA\ at $75^\circ$, b) the EW of the \mg~$2800$\AA\ line 
at $20^\circ$ and c) the EW of \la\ at $20^\circ$. The circles correspond 
to the simulation grid for two different values of $f_V$, and the fiducial 
model is marked with an orange star. 
We also show a higher X-ray luminosity model and a higher mass loss rate
with a red star.
}
\label{fig:grid}
\end{figure*}

@@

\bibliography{bibs/mybib.bib,bibs/stellar.bib,bibs/h14.bib,bibs/h13.bib,bibs/hamann.bib,bibs/krolik.bib}
